
---
title: "Perth Airport Rainfall Analysis \U0001F6EB"
output:
  html_document:
    toc: yes
    toc_depth: '2'
    toc_float:
      collapsed: no
    theme: darkly
    highlight: tango
    df_print: paged
    code_folding: hide
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Rainfall analysis and prediction for Perth Airport

Having recently moved to Perth from Newcastle I am intrigued by the different weather patterns here in Perth. The most striking feature is that rainfall in Perth is *seasonal* while in Newcastle it is mostly constant year round.

So I wanted to explore the nature of rainfall in Perth. Fortunately, it is easy to get data for any Australian weather site from the [Bureau of Meterology website](http://www.bom.gov.au/climate/data/?ref=ftr). Unfortunately, this no longer works for some reason (the element you click to download data still has a download tag, but instead of data, it returns a zipped 403 page). Instead, I have used the GHCN data for Perth Airport (which comes from the BOM anyway) which you can access [here](https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:ASN00009021/detail).

Documentation for the column names can be found [here](https://www.ncei.noaa.gov/pub/data/cdo/documentation/GHCND_documentation.pdf). In brief they are:
- STATION (17 characters) is the station identification code.  
- STATION_NAME (max 50 characters) is the name of the station.  
- GEOGRAPHIC_LOCATION (31 characters) is the latitude, longitude, and elevation above mean sea level.  
- DATE is the year of the record (4 digits) followed by month (2 digits) and day (2 digits).   
- PRCP = Precipitation (mm).  
- SNOW = Snowfall (mm).  
- SNWD = Snow depth (mm).  
- TMAX = Maximum temperature (Celsius).  
- TMIN = Minimum temperature (Celsius).  

The data I am using in this notebook is the daily rainfall totals and temperature for Perth Airport weather station. It has been continuously monitoring rainfall since May 1944! This allows us to get a good grasp on the seasonal rainfall patterns in Perth and also what the year to year trend looks like.

## code setup

Libraries:

```{r}
suppressMessages(library(data.table))
suppressMessages(library(tidyverse))
suppressMessages(library(splines))
suppressMessages(library(ggdark))
suppressMessages(library(patchwork))
require(gridExtra)
```

## import and clean data

Import data and add some useful columns to our rain dataframe. This includes a boolean for whether it rained. Really just allows me to have more comprehensible code and saves me adding an extra label to my ggplots.

```{r}
NOAA_data <- fread("airport_data/3184544.csv")
#filter for usability
#turns out the data attributes are useless lol
NOAA_data <- NOAA_data %>% select(!ends_with("_ATTRIBUTES"))
#add month and day cols for easier group-by summaries later
NOAA_data <- NOAA_data %>% 
  mutate(
    Year = format(DATE, "%Y") %>% as.numeric(),
    Month = month(DATE),
    Day = format(DATE, "%d") %>% as.numeric(),
    #bool for any rain
    did_rain = PRCP > 0,
    Month_name = month.name[Month] %>% fct_relevel(month.name)
  )
NOAA_data %>% head(2)
```

## Should you plan to go outside? Percent chance of any rain for Perth Airport

Probably the most striking feature of rain in Perth is just how seasonal it is. Looking at our first graph below we can see any given day in January has less than 10% chance of any rain while a given day in July has over 50% chance of any rain! The month to month change is dramatic, with the change in rainfall chance shifting rapidly in autumn and spring.

While summers are incredibly hot, and the UV index is dangerous, summer is a great time to plan outdoor activities that would be otherwise ruined by rain. In contrast, expect the opposite in winter.

```{r}
rainy_days <- NOAA_data %>% 
	group_by(Month, Day) %>% 
	summarise(
	  number_of_rainy_days = sum(did_rain), days_observed = n()
	) %>% 
	mutate(proportion_rained = number_of_rainy_days/days_observed)

rainy_days$index <- 1:length(rainy_days$Day)
rainy_days <- rainy_days %>% 
  mutate(date_str = str_c(Day, Month, sep = "/"))
custom_breaks = c(
	1,32,32+29,32+29+31,92+30,153,153+30,214,245,275,306,336,366
)
rainy_days %>% 
	ggplot(aes(index, proportion_rained*100)) +
	geom_point() +
	geom_smooth(span=0.5) +
	scale_y_continuous(breaks = seq(0,1,0.1)*100) +
	ylab("Percentage of days with any rain") +
	xlab("Date") +
	scale_x_continuous(
	  breaks = custom_breaks, labels = rainy_days$date_str[custom_breaks]
	) +
	ggtitle("Should you plan to go outside? Percent chance of any rain for Perth Airport") +
	dark_theme_classic()
```

## Changing rainfall with climate change

Perth is experiencing huge changes to it's rainfall. We can illustrate this in our next plot that shows the total rainfall each year. As we can there is steady trend of decreasing rainfall year-on-year.

```{r}
rainfall_totals_yearly <- NOAA_data %>% 
  group_by(Year) %>% summarise(total.Yearly.Rainfall = sum(PRCP)) 

rainfall_totals_yearly %>% 
  ggplot(aes(Year, total.Yearly.Rainfall)) +
  geom_point() +
  geom_smooth(method="lm") +
  #use expand in scale param to remove annoying scale padding
  scale_x_continuous(breaks = c(1945,1960,1980,2000,2020), expand = c(0.01,0.01)) +
  scale_y_continuous(
    expand = c(0.01,0.01), 
    limits = c(0,max(rainfall_totals_yearly$total.Yearly.Rainfall))
  ) +
  ggtitle("Total yearly rainfall for Perth Airport (1945 to 2022)") +
  dark_theme_classic()
```

To get some specific numbers on this, we can run a linear regression and consider the intercept and slope. This will produce the same result as the line through the graph above, as `ggplot::geom_smooth(method="lm")` is using the same function as below (`lm()`).

```{r}
#to use our intercept we set 1944 to be "year zero"
base_year = 1944
rainfall_totals_yearly <- rainfall_totals_yearly %>% mutate(Year_zeroed = Year-base_year)
rainfall_model_lm <- lm("total.Yearly.Rainfall ~ Year_zeroed" , data = rainfall_totals_yearly)
broom::tidy(summary(rainfall_model_lm))
```

We can see that in 1944, the average rainfall is 866mm (i.e. the intercept value) and that rainfall decreases at a rate of 2.8mm a year. In 2020 we would expect the rainfall to be `r round(predict(rainfall_model_lm, list(Year_zeroed = 2020-base_year)),2)` mm (the actual total was 636mm). That is a decrease of `r round(predict(rainfall_model_lm, list(Year_zeroed = 1945-base_year)),2)-round(predict(rainfall_model_lm, list(Year_zeroed = 2020-base_year)),2)`mm in just 75 years.

Here is a table showing some predicted rainfall values compared to actual totals, and then extended into the future. We can see if trends continued then our model predicts that rainfall would dwindle to zero. Of course, we can't expect trends to continue, and extending a relationship outside the measured boundaries is always 

```{r}
Years = c(1950, 1970, 2020, 2040, 2080, 2100, 2200)
df <- broom::tidy(predict(
  rainfall_model_lm, list(
    Year_zeroed=Years-base_year
  )
))
df$Year <- Years
df$Predicted_Rainfall <- df$x
df <- merge(df, rainfall_totals_yearly %>% 
  filter(Year %in% Years) %>% select(Year, total.Yearly.Rainfall) %>%
    rename(Actual_rainfall = total.Yearly.Rainfall),
  all.x = TRUE)
df %>% 
  select(Year, Predicted_Rainfall, Actual_rainfall) %>% 
  mutate(Difference = Actual_rainfall - Predicted_Rainfall)
```



## Changing rainfall, a deeper dive

This dataset is large enough for us to consider a deeper dive into Perth's changing rainfall. We can explore a few questions with graphs and statistical models:
1. in which months is rainfall increasing or decreasing?
2. is the time of peak rainfall each year changing or constant?

## in which months is rainfall increasing or decreasing?

It is interesting to consider whether rainfall patterns are changing in the same way for all months and times of year. We can use linear models to approach this in two different ways:

1. We can run linear models of `Year = total_rainfall` separately for each month
2. We can use a single linear model with month as an interaction term, i.e `Year = total_rainfall * month`

### linear models for each month

While the latter is more elegant, the former is easier to illustrate on a graph. We can see the results on the graph below ("Perth rainfall monthly totals"). Here, we can see that the relationship between Total Rainfall and Year differs across the year. We can also plot in on the winter and summer months to get a better feel for this.

We can see that rainfall totals are decreasing in winter, but appear to be steady or possibly increasing in summer (although we can't be certain of this as the p-values for the coefficients are not even nominally significance).

```{r, fig.align="center", fig.width=10, fig.height=15}
min_year = NOAA_data$Year %>% min
max_year = NOAA_data$Year %>% max
rainfall_totals <- NOAA_data %>% 
	group_by(Year, Month) %>% 
	summarise(total_rainfall = sum(PRCP))

allmonths <- rainfall_totals %>% 
	filter(Year != 1944) %>% 
	ggplot(aes(Year, total_rainfall)) + 
	facet_grid(cols=vars(Month)) +
	geom_point(alpha = 0.5) +
	geom_smooth(method="lm", colour = "red") +
	scale_x_continuous(breaks = c(min_year+1, max_year-1)) +
	ggtitle("Perth rainfall monthly totals for each month between 1945 and 2022") +
  dark_theme_classic() +
	theme(axis.text.x=element_text(angle=60,hjust=1))
winter <- rainfall_totals %>% 
	filter(Year != 1944, Month %in% 6:8) %>% 
	ggplot(aes(Year, total_rainfall)) + 
	facet_grid(cols=vars(Month)) +
	geom_point(alpha = 0.5) +
	geom_smooth(method="lm", colour = "red") +
	scale_x_continuous(breaks = c(min_year+1, max_year-1)) +
	ggtitle("Winter Months") +
  dark_theme_classic() +
	theme(axis.text.x=element_text(angle=60,hjust=1))
summer <- rainfall_totals %>% 
	filter(Year != 1944, Month %in% c(1:2,12)) %>% 
	ggplot(aes(Year, total_rainfall)) + 
	facet_grid(cols=vars(Month)) +
	geom_point(alpha = 0.5) +
	geom_smooth(method="lm", colour = "red") +
	scale_x_continuous(breaks = c(min_year+1, max_year-1)) +
	ggtitle("Summer months") +
  dark_theme_classic() +
	theme(axis.text.x=element_text(angle=60,hjust=1))
allmonths/(winter | summer)
```

We can inspect the validity of this observation by looking at the coefficients and p-values for the `Year ~ Rainfall` relationship for each month. The table and plot below show the coefficients, standard error estimates, and p-values for each month's model. While the p-values, for the coefficient, for almost all months were nominally significant (indicating that the relationship was not zero, i.e no change) only June was significant after correcting for multiple testing.

Interestingly, the *rate of change*, as expressed by the absolute magnitude of the coefficient differs considerably between months. The winter months had by far the largest rates of change. Since the winter months have negative coefficients, we interpret this as saying that rainfall is decreasing in these months much faster than in the warmer months.

We can also see the summer months have very little estimated change in their year to year rainfall totals. But there are tentative, although non-significant, indications of increasing rainfall.

```{r}
models.year_vs_rainfall_per_month <- rainfall_totals %>% 
	mutate(Year = Year - base_year) %>% 
	group_by(Month) %>% 
	group_modify(~ broom::tidy(lm(total_rainfall ~ Year, data = .)))
models.year_vs_rainfall_per_month %>% 
  mutate(nominal_significant = p.value < 0.05) %>% 
  mutate(passes_bonferroni = p.value < 0.05/12)
models.year_vs_rainfall_per_month %>% 
	filter(term == "Year") %>% 
	ggplot(aes(Month, estimate, ymin=estimate-std.error, ymax=estimate+std.error, colour = -log10(p.value))) + 
	dark_theme_classic() +
	geom_point() +
	geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 'dashed', alpha = 0.8, colour = "red") +
	ggtitle("Estimated change in average total monthly rainfall per year") +
	ylab("estimated change rainfall (mm)") +
	scale_x_continuous(breaks = 1:12)
```

### month as an interaction term

We can also briefly pause to see how using Month as an interaction term can robustly support the hypothesis that the year-on-year change in rainfall total is influenced by the month.

In order for this to work effectively we need to modify our Month variable so that it is either a `factor` or so that the winter months are the lowest or highest valued months.

While using month as a factor avoids any assumptions about months being in a specific order, it also dilutes our test by requiring each Month to act as it's own variable.

Instead, we can reparameterise our Month variable so that summer months have larger absolute values and winter months have smaller absolute values. We do this by subtracting the Month integer by 6 and taking the absolute value of this, leaving us with 7 unique values from 0 and 6.

```{r}
rainfall_totals <- rainfall_totals %>% 
  mutate(Year_zeroed = Year - base_year) %>% 
  mutate(Month_reparam = abs(Month-6))
year.by.month.model <- lm(total_rainfall ~ Year_zeroed*Month_reparam, data=rainfall_totals)
broom::tidy(summary(year.by.month.model))
```

We can now run our model using rainfall totals for each month, year pair. The result shows approximately what we would expect; rainfall is estimated to be decreasing year-on-year while there is a very strong negative relationship between our reparamterised Month variable (i.e rainfall decreases with increasing values, which fits with summer (larger values) having very low rainfall). We can also see that our interaction term indicates a small, but statistically significant, interaction between Year and Month. The coefficient is positive, which supports the argument that rainfall is decreasing in winter while the relationship is weaker in the warmer months.

## is the time of peak rainfall each year changing or constant?

Another apparent trend in Perth's weather pattern is that the period of peak rainfall is shifting to later in the year. Traditionally, June has been the month with the highest rainfall totals. But as we have seen above, June is experiencing a faster decline in rainfall with each year compared to any other month. This might lead us to think that another winter month might end up with greater rainfall in future years.

To test this in a visual way, we can compare 

```{r}
rainfall_totals_may_june_higher <- rainfall_totals %>% 
	filter(Month == 5 | Month == 6) %>% 
	#cols become Year june july
	select(Year, Month, total_rainfall) %>% 
	pivot_wider(names_from = c(Month), values_from = total_rainfall) %>% 
	mutate(higher_rain = case_when(
		`5` > `6` ~ "June",
		`6` > `5` ~ "July",
	)) %>% 
	mutate(higher_rain_ordinal = case_when(
		higher_rain == "May" ~ 0,
		higher_rain == "June" ~ 1
	)) %>% 
	mutate(delta = `6` - `5`)
rainfall_totals_may_june_higher %>% 
	ggplot(aes(Year, delta)) +
	geom_point() +
	geom_hline(yintercept = 0, linetype = 'dashed', alpha = 0.8, colour = "red") +
	geom_smooth(method="glm") +
	ggtitle("Difference in total rainfall in June compared to May\n(Perth Airport, BOM data)") +
	ylab("Difference (mm)") +
	dark_theme_bw() +
	ylim(-400,400) +
	annotate("text", label = "June", x = 1980, y = 300, size = 8) +
	annotate("text", label = "May", x = 1980, y = -300, size = 8)

rainfall_totals_june_july_higher <- rainfall_totals %>% 
	filter(Month > 5, Month < 8) %>% 
	#cols become Year june july
	select(Year, Month, total_rainfall) %>% 
	pivot_wider(names_from = c(Month), values_from = total_rainfall) %>% 
	mutate(higher_rain = case_when(
		`6` > `7` ~ "June",
		`7` > `6` ~ "July",
	)) %>% 
	mutate(higher_rain_ordinal = case_when(
		higher_rain == "June" ~ 0,
		higher_rain == "July" ~ 1
	)) %>% 
	mutate(delta = `7` - `6`)
rainfall_totals_june_july_higher %>% 
	ggplot(aes(Year, delta)) +
	geom_point() +
	geom_hline(yintercept = 0, linetype = 'dashed', alpha = 0.8, colour = "red") +
	geom_smooth(method="glm") +
	ggtitle("Difference in total rainfall in July compared to June\n(Perth Airport, BOM data)") +
	ylab("Difference (mm)") +
	dark_theme_bw() +
	ylim(-400,400) +
	annotate("text", label = "July", x = 1980, y = 300, size = 8) +
	annotate("text", label = "June", x = 1980, y = -300, size = 8)

rainfall_totals_july_august_higher <- rainfall_totals %>% 
	filter(Month == 7 | Month == 8) %>% 
	#cols become Year june july
	select(Year, Month, total_rainfall) %>% 
	pivot_wider(names_from = c(Month), values_from = total_rainfall) %>% 
	mutate(higher_rain = case_when(
		`7` > `8` ~ "July",
		`8` > `7` ~ "August"
	)) %>% 
	mutate(higher_rain_ordinal = case_when(
		higher_rain == "July" ~ 0,
		higher_rain == "August" ~ 1
	)) %>% 
	mutate(delta = `8` - `7`)

rainfall_totals_july_august_higher %>% 
	ggplot(aes(Year, delta)) +
	geom_point() +
	geom_hline(yintercept = 0, linetype = 'dashed', alpha = 0.8, colour = "red") +
	geom_smooth(method="glm") +
	ggtitle("Difference in total rainfall in August compared to July\n(Perth Airport, BOM data)") +
	ylab("Difference (mm)") +
	dark_theme_bw() +
	ylim(-400,400) +
	annotate("text", label = "August", x = 1980, y = 300, size = 8) +
	annotate("text", label = "July", x = 1980, y = -300, size = 8)
```



# Temperature trends at Perth Airport

The NOAA supplied dataset also includes daily min and max temperatures 

```{r}
temperature.yearly.avg <- NOAA_data %>% 
  group_by(Year) %>% 
  summarise(
    TMAX = mean(TMAX, na.rm = T),
    TMIN = mean(TMIN, na.rm = T),
    rain_total = sum(PRCP)
  ) 

temperature.yearly.avg %>% 
  pivot_longer(cols=c(TMIN, TMAX), names_to = "statistic", values_to = "temperature") %>% 
  ggplot(aes(Year, temperature, colour = statistic)) +
  geom_point() +
  geom_smooth(method = "lm") +
  dark_theme_bw() +
  ggtitle("Yearly min and max temperatures for Perth Airport (1944-2022)")
```

Monthly avg min and max
```{r}
temperature.month.by.year.avg <- NOAA_data %>% 
  group_by(Month, Month_name, Year) %>% 
  summarise(
    TMIN = mean(TMIN, na.rm = T),
    TMAX = mean(TMAX, na.rm = T),
    rain_total = sum(PRCP)
  ) 

temperature.month.by.year.avg %>% 
  filter(Year == 1945) %>% 
  ggplot(aes(x=Month, ymin=TMIN, ymax=TMAX)) +
  geom_linerange() +
  dark_theme_bw()
temperature.month.by.year.avg %>% 
  filter(Month == 1) %>% 
  ggplot(aes(x=Year, ymin=TMIN, ymax=TMAX)) +
  geom_linerange() +
  dark_theme_bw()
temperature.month.by.year.avg %>% 
  filter(Month == 7) %>% 
  ggplot(aes(x=Year, ymin=TMIN, ymax=TMAX)) +
  geom_linerange() +
  dark_theme_bw()
```

Does the difference between min and max stay the same? No, but the change we see is fairly small. It is maybe more interesting to note that it depends what month you are talking about.

```{r, fig.align="center", fig.width=10, fig.height=10}
#here we generate two plots and then present them together using patchwork

monthly.temp.trend.graph <- temperature.month.by.year.avg %>% 
  mutate(temp_range = TMAX-TMIN) %>% 
  #adjust the month names
  ggplot(aes(Year, temp_range)) +
  geom_point(alpha=0.2) +
  geom_smooth(method="lm") +
  facet_grid(cols = vars(Month_name)) +
  ggtitle("") +
  scale_x_continuous(breaks=c(1945,2022)) +
  dark_theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

models.temperature.month.by.year.avg <- temperature.month.by.year.avg %>% 
  mutate(
    temp_range = TMAX-TMIN,
    Year = Year - base_year
  ) %>% 
	group_by(Month) %>% 
	group_modify(~ broom::tidy(lm(temp_range ~ Year, data = .))) %>% 
  mutate(Month_name = month.name[Month] %>% fct_relevel(month.name))

graph.of.betas.temperature <- models.temperature.month.by.year.avg %>% 
  filter(term == "Year") %>% 
  ggplot(aes(Month_name, estimate, ymin=estimate-std.error, ymax=estimate+std.error, colour = -log10(p.value))) + 
	dark_theme_classic() +
	geom_point() +
	geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 'dashed', alpha = 0.8, colour = "red") +
  ggtitle("Estimated trend in difference between minimum and maxium temperature for each month") +
  ylab("mean change per year in degrees celsius")

monthly.temp.trend.graph / graph.of.betas.temperature
```



```{r}
temperature.month.by.year.avg %>% 
  filter(Month == 1) %>% 
  ggplot(aes(Year, TMAX)) +
  geom_point() +
  geom_smooth(method = "lm") +
  dark_theme_bw()
```

```{r}

```









## Temperature's correlation with rainfall?

Does temperature correlate with rainfall?

We have to consider Year in this model as it also correlates with both variables and is plausibly causal, i.e. a confounder.

We appear to get unstable estimates when including or excluding interaction... I guess because I also need to consider month?

```{r}
lm(rain_total ~ Year*TMAX, data = temperature.yearly.avg) %>%
    summary %>% broom::tidy() %>% 
    mutate(across(c("estimate", "std.error", "statistic", "p.value"), round, 2))
lm(rain_total ~ Year+TMAX, data = temperature.yearly.avg) %>%
    summary %>% broom::tidy() %>% 
    mutate(across(c("estimate", "std.error", "statistic", "p.value"), round, 2))
```

Monthly model:

The monthly model makes a lot more sense. But maybe that is partly due to increased sample size?

```{r}
lm(rain_total ~ Year*TMAX*Month, data = temperature.month.by.year.avg) %>% summary
```

I could look at a single month, say July:

```{r}
temperature.month.by.year.avg %>% 
    filter(Month == 7) %>% 
    lm(rain_total ~ TMAX, data =.) %>% 
    summary
```
```{r}
temperature.month.by.year.avg %>% 
    filter(Month == 4) %>% 
    ggplot(aes(TMAX, rain_total, colour=Year)) +
    geom_point() +
    geom_smooth(method="lm") +
    dark_theme_bw()
```


I don't think there is an easy way to untangle the relationships between temperature, year, and rainfall. Or at least I am not seeing it here.


# BONUS: Perth Airport Elevation

```{r}
NOAA_data %>% 
  ggplot(aes(DATE, ELEVATION)) +
  geom_point() +
  geom_line() +
  dark_theme_bw() +
  ggtitle("Elevation of Perth Airport between 1944 and 2022")
```


