#test

```{r}
suppressMessages(library(data.table))
suppressMessages(library(tidyverse))
```

Import NOAA data

Download from here: https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:ASN00009021/detail

Relevant column specs from documentation: https://www.ncei.noaa.gov/pub/data/cdo/documentation/GHCND_documentation.pdf

STATION (17 characters) is the station identification code. Please see http://www1.ncdc.noaa.gov/pub/data/ghcn/daily/ghcnd-stations.txt
STATION_NAME (max 50 characters) is the name of the station (usually city/airport name). Optional output field.
GEOGRAPHIC_LOCATION (31 characters) is the latitude (decimated degrees w/northern hemisphere
    values > 0, southern hemisphere values < 0), longitude (decimated degrees w/western hemisphere
    values < 0, eastern hemisphere values > 0) and elevation above mean sea level (tenths of meters). This is
    an optional output field.
DATE is the year of the record (4 digits) followed by month (2 digits) and day (2 digits).
PRCP = Precipitation (mm or inches as per user preference, inches to hundredths on Daily Form pdf file)
SNOW = Snowfall (mm or inches as per user preference, inches to tenths on Daily Form pdf file)
SNWD = Snow depth (mm or inches as per user preference, inches on Daily Form pdf file)
TMAX = Maximum temperature (Fahrenheit or Celsius as per user preference, Fahrenheit to tenths on Daily Form pdf file
TMIN = Minimum temperature (Fahrenheit or Celsius as per user preference, Fahrenheit to tenths on Daily Form pdf file

```{r}
NOAA_data <- fread("airport_data/3184544.csv")
#filter for usability
#turns out the data attributes are useless lol
NOAA_data <- NOAA_data %>% select(!ends_with("_ATTRIBUTES"))
```

```{r}
NOAA_data %>% summary
```

Snow depth data?!

Looks implausible as the min temp was 12C. Rainfall total seems reasonable though. Also a bit alarmed by TMIN > TAVG which should not be true under any circumstance?
```{r}
NOAA_data %>% 
  filter(!is.na(SNWD)) %>% 
  select(!ends_with("_ATTRIBUTES"))
```
Or did it snow the night before? No.
```{r}
NOAA_data %>% filter(DATE == as.Date("1999-09-26"))
```

